Imputation of Incomplete Multilevel Data with R

### 2023-12-12 docs/template/template.tex
Template manuscript

\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Imputation of Incomplete Multilevel Data with R},
  pdfauthor={Hanne I. Oberman; Johanna Muñoz; Valentijn M.T. de Jong; Gerko Vink; Thomas P.A. Debray},
  pdfkeywords={missing data, multilevel, clustering, mice, R},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

%% an abstract and keywords
\Abstract{This tutorial illustrates the imputation of incomplete
multilevel data with the \proglang{R} packackage \pkg{mice}. Our scope
is only simple multilevel models, to show how imputation can yield less
biased estimates from incomplete clustered data. More complex models can
be accomodated, but are outside the scope of this paper. Incomplete
multilevel data requires careful consideration of the missing data
problem and analysis strategy. In this tutorial, we focus on a popular
strategy for accommodating missingness in multilevel data: replacing the
missing data with one or more plausible values, i.e.,
imputation.Imputation separates the missing data problem from the main
analysis and the completed data can be analyzed as if it has been fully
observed. This tutorial illustrates the imputation of incomplete
multilevel data with the statistical programming language R. We aim to
show how imputation can yield less biased estimates from incomplete
clustered data. We provide practical guidelines and code snippets for
different missing data situations, including non-ignorable missingness
mechanisms. For brevity, we focus on multilevel imputation using chained
equations with the R mice package and its adjacent packages.}

%% at least one keyword must be supplied
\Keywords{missing
data, multilevel, clustering, \pkg{mice}, \proglang{R}}
\Plainkeywords{missing data, multilevel, clustering, mice, R}

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Hanne I. Oberman\\
Methodology and Statistics\\
Padualaan 14\\
Utrecht The Netherlands\\
E-mail: \email{h.i.oberman@uu.nl}\\
URL: \url{https://www.hanneoberman.github.io}\\
\\~
Johanna Muñoz\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~
Valentijn M.T. de Jong\\
Julius Centre for Health Sciences and Primary Care\\
Utrecht The Netherlands\\
\\~
Gerko Vink\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~
Thomas P.A. Debray\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~

@@

\begin{document}
\maketitle
\hypertarget{sec-intro}{%
\section{Introduction: Clustering and incomplete data}\label{sec-intro}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  missing data occur often in data with human subjects
\item
  missing data may be resolved, but need to be handled in accordance
  with the analysis of scientific interest
\item
  in human-subjects research, there is often clustering, which may be
  captured with multilevel modeling techniques
\item
  if the analysis of scientific interest is a multilevel model, the
  missing data handling method should accommodate the multilevel
  structure of the data
\item
  both missingness and multilevel structures require advanced
  statistical techniques
\item
  this tutorial sets out to facilitate empirical researchers in
  accommodating both multilevel structures as well as missing data.
\item
  we illustrate the use of the software by means of three case studies
  from the social and biomedical sciences.
\end{enumerate}

In hierarchical datasets, clustering is a concern because the
homoscedasticity in the error terms cannot be assumed across clusters
and the relationship among variables may vary at different hierarchical
levels. When multiple imputation is used to deal with missing data, as
the imputation and analysis process is performed separately, it is
necessary that imputation model being congenial with the main analysis
model (Meng, 1994), e.g.~if the main model accounts for the hierarchical
structure also imputation model should do it (Audigier, 2021). Not
including clustering into the imputation process may lead to effect
estimates with smaller standard errors and inflated type I error.

\hypertarget{overview-of-software}{%
\subsection{overview of software}\label{overview-of-software}}

The popular \pkg{mice} package in \proglang{R} \citet{R}

\hypertarget{scope}{%
\subsection{scope}\label{scope}}

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice} in \proglang{R}. \pkg{mice} has become the de-facto
standard for imputation by chained equations, which iteratively solves
the missingness on a variable-by-variable basis. \pkg{mice} is known to
yield valid inferences under many different missing data circumstances
\citep{buur18}.

We provide practical guidelines and code snippets for different missing
data situations, including non-ignorable mechanisms. For reasons of
brevity, we focus on multilevel imputation by chained equations with
\pkg{mice} exclusively; other imputation methods and packages \citep[see
e.g.][ and \citet{grun18}]{audi18} are outside the scope of this
tutorial. Assumed knowledge includes basic familiarity with the
\pkg{lme4} notation for multilevel models (see Table \ref{tab:mod}).

We illustrate imputation of incomplete multilevel data using three case
studies:

\begin{itemize}
\tightlist
\item
  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
  with data that are MAR,][]{mice};
\item
  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
  studies with data that are MAR,][]{metamisc};
\item
  \texttt{obesity} from the \pkg{micemd} package {[}simulated data on
  obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
  that are MNAR{]}.
\end{itemize}

For each of these datasets, we discuss the nature of the missingness,
choose one or more imputation models and evaluate the imputed data, but
we will also highlight one specific aspect of the imputation workflow.

This tutorial is dedicated to readers who are unfamiliar with multiple
imputation. More experienced readers can skip the introduction (case
study 1) and directly head to practical applications of multilevel
imputation under MAR conditions (case study 2) or under MNAR conditions
(case study 3).

\hypertarget{sec-models}{%
\section{Background}\label{sec-models}}

\hypertarget{concepts-in-multilevel-data}{%
\subsection{concepts in multilevel
data}\label{concepts-in-multilevel-data}}

Many datasets include individuals that are clustered together, for
example in geographic regions, or even different studies. In the
simplest case, individuals (e.g., students) are nested within a single
cluster (e.g., school classes). More complex clustered structures may
occur when there are multiple hierarchical levels (e.g., students in
different schools or patients within hospitals within regions across
countries), or when the clustering is non-nested (e.g., electronic
health record data from diverse settings and populations within large
databases). With clustered data we generally assume that individuals
from the same cluster tend to be more similar than individuals from
other clusters. In statistical terms, this implies that observations
from the same cluster are not independent and may in fact be correlated.
If this correlation is left unaddressed, estimates of \emph{p} values,
confidence intervals even model parameters are prone to bias
\citep{loca01}. Statistical methods for clustered data typically adopt
hierarchical models that explicitly describe the grouping of
observations. These models are also known as `multilevel models',
`hierarchical models', `mixed effect models', `random effect models',
and in the context of time-to-event data as `frailty models'. Table
\ref{tab:clus} provides an overview of some key concepts in multilevel
modeling.

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, arc=.35mm, rightrule=.15mm, breakable, opacityback=0, bottomrule=.15mm, colback=white, leftrule=.75mm, left=2mm]

Box 1. The intraclass correlation coefficient.

@@

In \proglang{R}, multilevel models may be fitted using the package
\pkg{lme4}. For linear mixed-effects models, the function

\begin{verbatim}
lmer(formula, data, ...)
\end{verbatim}

\hypertarget{concepts-in-missing-data}{%
\subsection{concepts in missing data}\label{concepts-in-missing-data}}

missing data mechanisms etc.

As with any other dataset, clustered datasets may be impacted by
missingness in much the same way. Several strategies can be used to
handle missing data, including complete case analysis and imputation. We
focus on the latter approach and discuss statistical methods for
replacing the missing data with one or more plausible values. Imputation
separates the missing data problem from the analysis and the completed
data can be analyzed as if it were completely observed. It is generally
recommended to impute the missing values more than once to preserve
uncertainty due to missingness and to allow for valid inferences (c.f.
Rubin 1976).

With incomplete clustered datasets we can distinguish between two types
of missing data: sporadic missingness and systematic missingness
\citep{resc13}. Sporadic missingness arises when variables are missing
for some but not all of the units in a cluster \citep{buur18, jola18}.
For example, it is possible that test results are missing for several
students in one or more classes. When all observations are missing
within one or more clusters, data are said to be systematically missing.
Sporadic missingness is visualized in Figure XYZ.

\begin{figure}[h]

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-3-1.pdf}

@@

@@

Column \(X_1\) in Figure 1 is completely observed, column \(X_2\) is
systematically missing in cluster 2, and column \(X_3\) is sporadically
missing. To analyze these incomplete data, we have to take the nature of
the missingness and the cluster structure into account. For example, the
sporadic missingness in \(X_3\) could be easily amended if this would be
a cluster-level variable (and thus constant within clusters). We could
then just extrapolate the true (but missing) value of \(X_3\) for unit 1
from unit 2, and the value for unit 4 from unit 3. If \(X_3\) would
instead be a unit-level variable (which may vary within clusters), we
could not just recover the unobserved `truth', but would need to use
some kind of missing data method, or discard the incomplete units
altogether (i.e., complete case analysis). Complete case analysis can
however introduce bias in statistical inferences and lowers statistical
power. Further, with the systematic missingness in \(X_2\), it would be
impossible to fit a multilevel model without accommodating the
missingness in some way. Complete case analysis in that case would mean
excluding the entire cluster from the analyses. The wrong choice of
missing data handling method can thus be extremely harmful to the
inferences.

Imputation of missing data requires consideration of the mechanism
behind the missingness. Rubin proposed to distinguish between data that
are missing completely at random (MCAR), data that are missing at random
(MAR) and data that are missing not at random (MNAR; see Table
\ref{tab:miss}). For each of these three missingness generating
mechanisms, different imputation strategies are warranted
(\citet{yuce08} and \citet{hox15}). We here consider the general case
that data are MAR, and expand on certain MNAR situations.

\hypertarget{imputation-with-mice}{%
\subsection{imputation with mice}\label{imputation-with-mice}}

The \proglang{R} package \pkg{mice} provides a framework for imputing
incomplete data on a variable-by-variable basis. The \fct{mice} function
allows users to flexibly specify how many times and under what model the
missing data should be imputed. This is reflected in the first four
function arguments

\begin{verbatim}
mice(data, m, method, predictorMatrix, ...)
\end{verbatim}

where \texttt{data} refers to the incomplete dataset, \texttt{m}
determines the number of imputations, \texttt{method} denotes the
functional form of the imputation model and \texttt{predictorMatrix}
specifies the interrelational dependencies between variables and
imputation models (i.e., the set of predictors to be used for imputing
each incomplete variable).

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, arc=.35mm, rightrule=.15mm, breakable, opacityback=0, bottomrule=.15mm, colback=white, leftrule=.75mm, left=2mm]

Box 2. The \texttt{methods}.

@@

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, arc=.35mm, rightrule=.15mm, breakable, opacityback=0, bottomrule=.15mm, colback=white, leftrule=.75mm, left=2mm]

Box 3. The predictor matrix. The entries corresponding to the level-1
predictors are coded with a 3, indicating that both the original values
as well as the cluster means of the predictor are included into the
imputation model. The entry of 4 in the predictor matrix adds three
variables to the imputation model for the imputation model predictor:
the value of the predictor, the cluster means of the predictor and the
random slopes of the predictor. - -2 = cluster variable - 1 = overall
effect - 3 = overall + group-level effect - 4 = individual-level
(random) and group-level (fixed) effect

@@

\hypertarget{sec-workflow}{%
\section{Multilevel imputation workflow}\label{sec-workflow}}

There are different strategies that can be adopted in the imputation
process that account for clustering: inclusion of cluster indicator
variable, performing a separate imputation process for each cluster, or
performing a simultaneous imputation process by using an imputation
method that accounts for clustering.(Stata:
https://www.stata.com/support/faqs/statistics/clustering-and-mi-impute/)
TODO: replace ref.

The selection of each strategy depends mainly on the assumptions in the
main analysis and also on the restriction of the analyzed data.

Regarding the restrictions imposed by the data, for instance, the use of
cluster indicator variables is restricted in datasets where there are
not many clusters and many observations per cluster (Graham, 2009). The
last restriction is also required when imputations are performed on each
cluster separately. When this restriction cannot be achieved, one can
use an imputation model that simultaneously imputes all clusters using a
hierarchical model (Allison 2002).

Under this hierarchical imputation model, observations within clusters
are correlated and this correlation is modeled by a random effect so the
hierarchical model can be estimated even when there are few observations
per cluster. However, this strategy is best suited for balanced data
(Grund, 2017) and when random effects model is appropriated, i.e.~the
number of clusters is adequate. (Austin,2018).

Here it is important to evaluate the assumptions imposed by the main
model, for instance by using the cluster indicator strategy may lead to
bias estimates when the model is based on a hierarchical model
(Taaljard,2008). Even when an imputation strategy congenial with the
main model is preferred, it is important to consider whether it is
appropriate for the data as a less complex imputation strategies may
also lead to unbiased estimates in certain scenarios(Bailey 2020). For
instance, in causal effect analysis, separately imputation may lead to
smaller bias when the size of the smaller exposure cluster is large,
compared with an imputation model that includes exposure-confounder
interactions. (Zhang,2023).

Below we provide a imputation workflow that can be used in general to
impute cluster data.

>>>>

\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Imputation of Incomplete Multilevel Data with R},
  pdfauthor={Hanne I. Oberman; Johanna Muñoz; Valentijn M.T. de Jong; Gerko Vink; Thomas P.A. Debray},
  pdfkeywords={JSS, style guide, comma-separated, not capitalized, R},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

%% an abstract and keywords
\Abstract{This short article illustrates how to write a manuscript for
the \emph{Journal of Statistical Software} (JSS) using its {\LaTeX}
style files. Generally, we ask to follow JSS's style guide and FAQs
precisely. Also, it is recommended to keep the {\LaTeX} code as simple
as possible, i.e., avoid inclusion of packages/commands that are not
necessary. For outlining the typical structure of a JSS article some
brief text snippets are employed that have been inspired by
\citet{ZeileisKleiberJackman2008}, discussing count data regression in
\proglang{R}. Editorial comments and instructions are marked by vertical
bars.}

%% at least one keyword must be supplied
\Keywords{JSS, style guide, comma-separated, not
capitalized, \proglang{R}}
\Plainkeywords{JSS, style guide, comma-separated, not capitalized, R}

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Hanne I. Oberman\\
Methodology and Statistics\\
Padualaan 14\\
Utrecht The Netherlands\\
E-mail: \email{Achim.Zeileis@R-project.org}\\
URL: \url{https://www.zeileis.org/}\\
\\~
Johanna Muñoz\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~
Valentijn M.T. de Jong\\
Julius Centre for Health Sciences and Primary Care\\
Utrecht The Netherlands\\
\\~
Gerko Vink\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~
Thomas P.A. Debray\\
Julius Centre for Health Sciences and Primary Care\\
Universiteitsweg 100\\
Utrecht The Netherlands\\
\\~

@@

\begin{document}
\maketitle
\hypertarget{sec-intro}{%
\section{Introduction: Count data regression in R}\label{sec-intro}}

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]

The introduction is in principle ``as usual''. However, it should
usually embed both the implemented \emph{methods} and the
\emph{software} into the respective relevant literature. For the latter
both competing and complementary software should be discussed (within
the same software environment and beyond), bringing out relative
(dis)advantages. All software mentioned should be properly
\texttt{@cited}'d.~(See also \protect\hyperlink{sec-bibtex}{Using
BibTeX} for more details on \textsc{Bib}{\TeX}.)

For writing about software JSS requires authors to use the markup
\texttt{{[}{]}\{.proglang\}} (programming languages and large
programmable systems), \texttt{{[}{]}\{.pkg\}} (software packages), back
ticks like `code` for code (functions, commands, arguments, etc.).

If there is such markup in (sub)section titles (as above), a plain text
version has to be provided in the {\LaTeX} command as well. Below we
also illustrate how abbrevations should be introduced and citation
commands can be employed. See the {\LaTeX} code for more details.

@@

Modeling count variables is a common task in economics and the social
sciences. The classical Poisson regression model for count data is often
of limited use in these disciplines because empirical count data sets
typically exhibit overdispersion and/or an excess number of zeros. The
former issue can be addressed by extending the plain Poisson regression
model in various directions: e.g., using sandwich covariances or
estimating an additional dispersion parameter (in a so-called
quasi-Poisson model). Another more formal way is to use a negative
binomial (NB) regression. All of these models belong to the family of
generalized linear models (GLMs). However, although these models
typically can capture overdispersion rather well, they are in many
applications not sufficient for modeling excess zeros. Since
\citet{Mullahy1986} there is increased interest in zero-augmented models
that address this issue by a second model component capturing zero
counts. An overview of count data models in econometrics, including
hurdle and zero-inflated models, is provided in
\citet{CameronTrivedi2013}.

In \proglang{R} \citet{R}, GLMs are provided by the model fitting
functions \fct{glm} in the \pkg{stats} package and \fct{glm.nb} in the
\pkg{MASS} package \citep{VenablesRipley2002} along with associated
methods for diagnostics and inference. The manuscript that this document
is based on \citep{ZeileisKleiberJackman2008} then introduced hurdle and
zero-inflated count models in the functions \fct{hurdle} and
\fct{zeroinfl} in the \pkg{pcsl} package \citet{Jackman2015}. Of course,
much more software could be discussed here, including (but not limited
to) generalized additive models for count data as available in the
\proglang{R} packages \pkg{mgcv} \citet{Wood2006}, \pkg{gamlss}
\citet{StasinopoulosRigby2007}, or \pkg{VGAM} \citet{Yee2009}.

\hypertarget{sec-models}{%
\section{Models and software}\label{sec-models}}

The basic Poisson regression model for count data is a special case of
the GLM framework \citet{McCullaghNelder1989}. It describes the
dependence of a count response variable \(y_i\) (\(i = 1, \dots, n\)) by
assuming a Poisson distribution \(y_i \sim \mathrm{Pois}(\mu_i)\). The
dependence of the conditional mean \(E[y_i \, | \, x_i] = \mu_i\) on the
regressors \(x_i\) is then specified via a log link and a linear
predictor

\begin{equation}\protect\hypertarget{eq-mean}{}{
\log(\mu_i) \quad = \quad x_i^\top \beta,
}\label{eq-mean}\end{equation}

where the regression coefficients \(\beta\) are estimated by maximum
likelihood (ML) using the iterative weighted least squares (IWLS)
algorithm.

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]

TODO: Note that around the equation above there should be no spaces
(avoided in the {\LaTeX} code by \texttt{\%} lines) so that ``normal''
spacing is used and not a new paragraph started.

\proglang{R} provides a very flexible implementation of the general GLM
framework in the function \fct{glm} \citet{ChambersHastie1992} in the
\pkg{stats} package. Its most important arguments are

\begin{verbatim}
glm(formula, data, subset, na.action, weights, offset,
  family = gaussian, start = NULL, control = glm.control(...),
  model = TRUE, y = TRUE, x = FALSE, ...)
\end{verbatim}

where \texttt{formula} plus \texttt{data} is the now standard way of
specifying regression relationships in \proglang{R}/\proglang{S}
introduced in \citet{ChambersHastie1992}. The remaining arguments in the
first line (\texttt{subset}, \texttt{na.action}, \texttt{weights}, and
\texttt{offset}) are also standard for setting up formula-based
regression models in \proglang{R}/\proglang{S}. The arguments in the
second line control aspects specific to GLMs while the arguments in the
last line specify which components are returned in the fitted model
object (of class \class{glm} which inherits from \class{lm}). For
further arguments to \fct{glm} (including alternative specifications of
starting values) see \texttt{?glm}. For estimating a Poisson model
\texttt{family\ =\ poisson} has to be specified.

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]

As the synopsis above is a code listing that is not meant to be
executed, one can use either the dedicated \texttt{\{Code\}} environment
or a simple \texttt{\{verbatim\}} environment for this. Again, spaces
before and after should be avoided.

Finally, there might be a reference to a \texttt{\{table\}} such as
Table~\ref{tbl-overview}. Usually, these are placed at the top of the
page (\texttt{{[}t!{]}}), centered (\texttt{\textbackslash{}centering}),
with a caption below the table, column headers and captions in sentence
style, and if possible avoiding vertical lines.

\hypertarget{tbl-overview}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1216}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.6351}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Distribution
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Method
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Distribution
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Method
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Description
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
GLM & Poisson & ML & Poisson regression: classical GLM, estimated by
maximum likelihood (ML) \\
& & Quasi & ``Quasi-Poisson regression'\,': same mean function,
estimated by quasi-ML (QML) or equivalently generalized estimating
equations (GEE), inference adjustment via estimated dispersion
parameter \\
& & Adjusted & ``Adjusted Poisson regression'\,': same mean function,
estimated by QML/GEE, inference adjustment via sandwich covariances \\
& NB & ML & NB regression: extended GLM, estimated by ML including
additional shape parameter \\
Zero-augmented & Poisson & ML & Zero-inflated Poisson (ZIP), hurdle
Poisson \\
& NB & ML & Zero-inflated NB (ZINB), hurdle NB \\
\caption{\label{tbl-overview}Overview of various count regression
models. The table is usually placed at the top of the page
(\texttt{{[}t!{]}}), centered (\texttt{centering}), has a caption below
the table, column headers and captions are in sentence style, and if
possible vertical lines should be avoided.}\tabularnewline
\end{longtable}

For a simple illustration of basic Poisson and NB count regression the
\texttt{quine} data from the \pkg{MASS} package is used. This provides
the number of \texttt{Days} that children were absent from school in
Australia in a particular year, along with several covariates that can
be employed as regressors. The data can be loaded by

\begin{verbatim}
R> data("quine", package = "MASS")
\end{verbatim}

and a basic frequency distribution of the response variable is displayed
in Figure~\ref{fig-quine}.

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]

For code input and output, the style files provide dedicated
environments. Either the ``agnostic'' \texttt{\{CodeInput\}} and
\texttt{\{CodeOutput\}} can be used or, equivalently, the environments
\texttt{\{Sinput\}} and \texttt{\{Soutput\}} as produced by \fct{Sweave}
or \pkg{knitr} when using the \texttt{render\_sweave()} hook. Please
make sure that all code is properly spaced, e.g., using
\texttt{y\ =\ a\ +\ b\ *\ x} and \emph{not} \texttt{y=a+b*x}. Moreover,
code input should use ``the usual'' command prompt in the respective
software system. For \proglang{R} code, the prompt
\texttt{R\textgreater{}} should be used with \texttt{+} as the
continuation prompt. Generally, comments within the code chunks should
be avoided -- and made in the regular {\LaTeX} text instead. Finally,
empty lines before and after code input/output should be avoided (see
above).

\begin{figure}

{\centering

\includegraphics{article-visualization.pdf}

@@

\caption{\label{fig-quine}Frequency distribution for number of days
absent from school.}

@@

As a first model for the \texttt{quine} data, we fit the basic Poisson
regression model. (Note that JSS prefers when the second line of code is
indented by two spaces.)

\begin{verbatim}
R> m_pois <- glm(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine, family = poisson)
\end{verbatim}

To account for potential overdispersion we also consider a negative
binomial GLM.

\begin{verbatim}
R> library("MASS")
R> m_nbin <- glm.nb(Days ~ (Eth + Sex + Age + Lrn)^2, data = quine)
\end{verbatim}

In a comparison with the BIC the latter model is clearly preferred.

\begin{verbatim}
R> library("MASS")
R> BIC(m_pois, m_nbin)
\end{verbatim}

\begin{verbatim}
       df      BIC
m_pois 18 2046.851
m_nbin 19 1157.235
\end{verbatim}

Hence, the full summary of that model is shown below.

\begin{verbatim}
R> summary(m_nbin)
\end{verbatim}

\begin{verbatim}

Call:
glm.nb(formula = Days ~ (Eth + Sex + Age + Lrn)^2, data = quine, 
    init.theta = 1.60364105, link = log)

Coefficients: (1 not defined because of singularities)
            Estimate Std. Error z value Pr(>|z|)    
(Intercept)  3.00155    0.33709   8.904  < 2e-16 ***
EthN        -0.24591    0.39135  -0.628  0.52977    
SexM        -0.77181    0.38021  -2.030  0.04236 *  
AgeF1       -0.02546    0.41615  -0.061  0.95121    
AgeF2       -0.54884    0.54393  -1.009  0.31296    
AgeF3       -0.25735    0.40558  -0.635  0.52574    
LrnSL        0.38919    0.48421   0.804  0.42153    
EthN:SexM    0.36240    0.29430   1.231  0.21818    
EthN:AgeF1  -0.70000    0.43646  -1.604  0.10876    
EthN:AgeF2  -1.23283    0.42962  -2.870  0.00411 ** 
EthN:AgeF3   0.04721    0.44883   0.105  0.91622    
EthN:LrnSL   0.06847    0.34040   0.201  0.84059    
SexM:AgeF1   0.02257    0.47360   0.048  0.96198    
SexM:AgeF2   1.55330    0.51325   3.026  0.00247 ** 
SexM:AgeF3   1.25227    0.45539   2.750  0.00596 ** 
SexM:LrnSL   0.07187    0.40805   0.176  0.86019    
AgeF1:LrnSL -0.43101    0.47948  -0.899  0.36870    
AgeF2:LrnSL  0.52074    0.48567   1.072  0.28363    
AgeF3:LrnSL       NA         NA      NA       NA    
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for Negative Binomial(1.6036) family taken to be 1)

    Null deviance: 235.23  on 145  degrees of freedom
Residual deviance: 167.53  on 128  degrees of freedom
AIC: 1100.5

Number of Fisher Scoring iterations: 1

              Theta:  1.604 
          Std. Err.:  0.214

 2 x log-likelihood:  -1062.546 
\end{verbatim}

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]

As usual\ldots{}

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toprule=.15mm, left=2mm, breakable, arc=.35mm, colback=white, opacityback=0, rightrule=.15mm, leftrule=.75mm]

If necessary or useful, information about certain computational details
such as version numbers, operating systems, or compilers could be
included in an unnumbered section. Also, auxiliary packages (say, for
visualizations, maps, tables, \ldots) that are not cited in the main
text can be credited here.

@@

The results in this paper were obtained using
\proglang{R}\textasciitilde3.4.1 with the
\pkg{MASS}\textasciitilde7.3.47 package. \proglang{R} itself and all
packages used are available from the Comprehensive \proglang{R} Archive
Network (CRAN) at {[}https://CRAN.R-project.org/{]}.
